---
title: "Iowa Census Exploratory Analysis"
output: html_notebook
author: Nicole Wright
---
Goals of IA census Project:

1. getting more familiar with tidycensus package
2. finding underserved regions in iowa with respect to the variables of interest
3. producing content relevant to government and non-profit agencies in the state

What is an underserved populaitons have?
* Low income
* Homeless
* migrant farm workers
* High percentage under the Federal poverty level
* High percentage over 65 years old (how much, >20%)
* High infant mortality rates
* Low population to provide ratio

What content is relevant to government and non-profit agencies in Iowa?
* Underserved populations described above
* Statistics around children (poverty, education, stability of home and movement)

Use Iowa Census and ACS data to explore occupation, gender, income, and age statistics of Iowa. The goal of this exploratory analysis is to find intersting questions to ask about the data.This data may be used inconjuction with US hospital data at a later date.  


```{r Load Packages}
library(tidyverse)
library(tidycensus)
library(readr)
library(tmap)
library(tmaptools)
library(sf)
library(hydroTSM)
```


2010 US Census data, using the tidycensus package ( 1/1/2019).
Key: 6bfbcee4b4c37a9683ac9c0e1d174d5f0ad67674
pdf: https://cran.r-project.org/web/packages/tidycensus/tidycensus.pdf

```{r 2010 Census set up}
census_api_key("6bfbcee4b4c37a9683ac9c0e1d174d5f0ad67674", install = TRUE, overwrite = TRUE)
census_key <- "6bfbcee4b4c37a9683ac9c0e1d174d5f0ad67674"

# First time, reload your environment so you can use the key without restarting R.
readRenviron("~/.Renviron")
```
```{r Load Variables}
LVacs5 <- load_variables(2017, "acs5", cache = TRUE)
LVacs1 <- load_variables(2017, "acs1", cache = TRUE)
LVcn2010 <- load_variables(2010, "sf1", cache = TRUE)

#FAMILY INCOME IN THE PAST 12 MONTHS (IN 2017 INFLATION-ADJUSTED DOLLARS) :B10010_001
#OCCUPATION BY MEDIAN EARNINGS IN THE PAST 12 MONTHS (IN 2017 INFLATION-ADJUSTED DOLLARS) FOR THE CIVILIAN EMPLOYED POPULATION 16 YEARS AND OVER: B24011_030
#DETAILED OCCUPATION BY MEDIAN EARNINGS IN THE PAST 12 MONTHS (IN 2017 INFLATION-ADJUSTED DOLLARS) FOR THE FULL-TIME, YEAR-ROUND CIVILIAN EMPLOYED POPULATION 16 YEARS AND OVER: B24012_067


#PLACE OF BIRTH BY EDUCATIONAL ATTAINMENT IN THE UNITED STATES for 
  #No Highschool: B15003_002
  #Hischool: B06009_021
  #Associate's or some: B06009_022
  #Bachelor's: B06009_023
  #Graduates: B06009_024
#EDUCATIONAL ATTAINMENT BY PRESENCE OF A COMPUTER AND TYPES OF INTERNET SUBSCRIPTION IN HOUSEHOLD
  #Less High school, no computer: B28006_007
  #Less High schook, computer: B28006_003
  #Less High school, computer with dial up internet: B28006_004
  #Less High school, computer with broad band internet: B28006_005
#MEDIAN FAMILY INCOME FOR FAMILIES WITH GRANDPARENT HOUSEHOLDERS AND/OR SPOUSES LIVING WITH OWN GRANDCHILDREN UNDER 18 YEARS BY RESPONSIBILITY FOR OWN GRANDCHILDREN AND PRESENCE OF PARENT OF GRANDCHILDREN

  
```

Iowa appears to have a low percentage of Iowas per the total population is very low (<5% of Iowans are reported as farmers). the ERS 2016 study showes that only 49.2% of farm workers are US citizens.Could this be why report is so low?
https://www.ers.usda.gov/topics/farm-economy/farm-labor/

Make this section neater!!
```{r census_IA data}
#This is to get State data and geoid data for adding to actual data later
x0 <- get_decennial(geography = "county", variables = "P001001", year = 2010, key = census_key)

#2010 census variables, 
State <- unlist(strsplit(x0$NAME, split = ", ")) %>%
    matrix(ncol = 2, byrow = TRUE) %>% 
    as_tibble() %>%
    dplyr::rename(State = V2) %>%
    mutate(GEOID = x0$GEOID)

#2017 5 year acs
v01 <- c(total_population = "B00001_001", total_median_age = "B01002_001", med_family_income = "B19013_001", farming_income = "B24011_030", male_farmers_all = "C24010_031", female_farmers_all = "C24010_067", total_nonHigh = "B28006_002", nonHigh_no_computer = "B28006_007", nonHigh_computer = "B28006_003", nonHigh_computer_dialup = "B28006_004", nonHigh_computer_broadband = "B28006_005")
        

x01 <- get_acs(geography = "county", variables = v01, year = 2017, geometry = TRUE, cache_table = TRUE, survey = "acs5", shift_geo = TRUE, key = census_key) %>%
  left_join(State, by = "GEOID") 
  
acs_IA <- x01 %>%  
  filter(State == "Iowa") %>%
  select(-V1, -moe) %>%
  spread(variable, estimate) %>%
  mutate(total_farmers = male_farmers_all + female_farmers_all, 
         farmers_per = total_farmers/total_population*100,   
         farming_to_total_income = farming_income / med_family_income,
         nonHigh_no_computer_per = nonHigh_no_computer / total_nonHigh * 100,
         nonHigh_computer_per = nonHigh_computer / total_nonHigh * 100,
         nonHigh_computer_dialup_per = nonHigh_computer_dialup / total_nonHigh * 100, 
         nonHigh_computer_broadband_per = nonHigh_computer_broadband / total_nonHigh * 100) %>%
  select(County = NAME, total_population, total_median_age, med_family_income, total_farmers, farmers_per, farming_income, farming_to_total_income, contains("nonHigh"), State, GEOID, geometry) %>%
  mutate(rounded_income = round(med_family_income, -4), Age = cut_interval(total_median_age, n = 4, length = NULL, labels = c("25-30", "30-35", "35-40", ">40"))) %>%
  mutate(farmersf = cut_interval(farmers_per, n = 5, labels = c("<5", "5-10", "10-14", "14-18", ">18"))) %>%
   mutate(farmersi = cut_width(farming_to_total_income, width = .5))


```

Notes from above: cut_interval is used to discretize a numerical variable into distinct groups for factor analysis, a secondary varaibale is created representing how that variabe fits into the new labeled bin (Age: if 35 fall in the 30-35 bin). For the median family income variables and new rounded variable was created that would make more incomes match for factor analysis.

Non-highschool graduate variables note: recorded the total number of  people who did not graduate highschool, or eqivalent, (nonHigh_total) of that subset population recorded people who did not have access to computer, computer but no internet, computer with dial up internet, and computer with broadband internet. The idea is kids who have limited access to computers will have a higher rate of non-highschool grad. Obviously kids without computers or internet probably come from poor families where recourses are already low. Just interesting little tibit.

In the future I could creat multible data sets for exploring different subsets of iowa data.
----
Theme: I need a theme that has white background, no grid lines, ... basically completely blank and I would like to add legend title and shift or tilt the axis so that Iowa is level to the eye (maybe not nescessary..) Nadia reccommended  theme_minimal() with coord_sf(datum = NA)
Aslo pick universal color scheme... I don't really like what I've got sofar 061519


```{r theme}
theme(plot.background = element_rect(color = "white"), legend.title = element_blank()) 
theme_blank <- theme_minimal() + coord_sf(datum = NA)
#Error: Don't know how to add RHS to a theme object... fix
```


```{r Age data}
#Using ACS 2017 5-year survey data (same variables from 1 year)
##Population by sex and age 
##Yes, you need as.character(label)!!
LVacs5 <- load_variables(2017, "acs5", cache = TRUE)
View(LVacs5)

L2 <- LVacs5 %>%
  filter(concept == "SEX BY AGE") %>%
  as_tibble() %>%
  mutate(as.character(label)) %>%
  mutate(label = sapply(strsplit(label, split = 'Estimate!!Total!!', fixed = TRUE), function(x) (x[2]))) %>% 
  na.omit() %>%
  dplyr::select(name, label) 

v04 <- as.vector(L2$name) 
v05 <- as.vector(L2$label) 

x03 <- get_acs(geography = "county", 
               variables = v04, 
               state = "IA",
               year = 2017, 
               geometry = TRUE, 
               cache_table = TRUE, 
               survey = "acs5", 
               shift_geo = FALSE, 
               key = census_key)

Age_IA <- x03 %>%
  select(-moe) %>%
  spread(variable, estimate) %>%
  rename(County = NAME)

#Can't figure out how to include in the pipe.
names(Age_IA)[3:50] <- v05
colnames(Age_IA) <- gsub("!!", "_", colnames(Age_IA)) 
colnames(Age_IA) <- gsub(" ", "_", colnames(Age_IA)) 
```

```{r Total_Age_IA}
Total_Age_IA <- Age_IA %>% 
  mutate(Under_5_years = Male_Under_5_years + Female_Under_5_years,
         years_5to9 = Male_5_to_9_years + Female_5_to_9_years,
         years_10to14 = Male_10_to_14_years + Female_10_to_14_years,
         years_15to17 = Male_15_to_17_years + Female_15_to_17_years,
         Under_18_years = Under_5_years + years_5to9 + years_10to14 + years_15to17,
         years_18to24 = Male_18_and_19_years + Male_20_years + Male_21_years + Male_22_to_24_years +
            Female_18_and_19_years + Female_20_years + Female_21_years + Female_22_to_24_years,
         years_25to29 = Male_25_to_29_years + Female_25_to_29_years, 
         years_30to39 = Male_30_to_34_years + Male_35_to_39_years + Female_30_to_34_years + Female_35_to_39_years,
         years_40_49 = Male_40_to_44_years + Male_45_to_49_years +Female_40_to_44_years + Female_45_to_49_years,
         years_50to59 = Male_50_to_54_years + Male_55_to_59_years +Female_50_to_54_years + Female_55_to_59_years,
         years_60to64 = Male_60_and_61_years + Male_62_to_64_years + Female_60_and_61_years + Female_62_to_64_years,
         years_65to69 = Male_65_and_66_years + Male_67_to_69_years + Female_65_and_66_years + Female_67_to_69_years,
         years_70to74 = Male_70_to_74_years + Female_70_to_74_years,
         years_75to79 = Male_75_to_79_years + Female_75_to_79_years,
         years_80to84 = Male_80_to_84_years + Female_80_to_84_years,
         Over_85_years = Male_85_years_and_over + Female_85_years_and_over,
         Over_65_years = years_65to69 + years_70to74 + years_75to79 + years_80to84 + Over_85_years) 
Total_Age_IA[ , c(3:50)] <- NULL
View(Total_Age_IA)
```


```{r attempt at automation Age_IA}
#Can we automattially add male and female age groups together?
v06 <- names(Age_IA)[3:50] %>% as.vector()
  
v06 <- gsub("Male", " ", v06)
v06 <- gsub("Female", " ", v06)

v06 <-rm1stchar(v06, n = 2)
v06<- v06[-c(1:25)]  
v06 <- as.vector(v06)

gather(Age_IA, key = "Gender", value = "Age") 

```

```{r get IA_poverty}
#Using ACS 2017 5-year survey data (same variables from 1 year)
##Population by sex and age below the povery level in the last 12months.
LVacs1 <- load_variables(2017, "acs1", cache = TRUE)
L1 <- LVacs1 %>%
  filter(grepl('B17001_', name)) %>%
  filter(!grepl('above', label)) %>%
  as_tibble() %>%
  mutate(as.character(label)) %>%
  mutate(label = sapply(strsplit(label, split = '!!Income in the past 12 months below poverty level!!', fixed = TRUE), function(x) (x[2]))) %>% 
  na.omit() %>%
  dplyr::select(name, label)

v02 <- as.vector(L1$name)
v03 <- as.vector(L1$label) 

x02 <- get_acs(geography = "county", 
               variables = v02, 
               state = "IA",
               year = 2017, 
               geometry = TRUE, 
               cache_table = TRUE, 
               survey = "acs5", 
               shift_geo = FALSE, 
               key = census_key)

poverty_IA <- x02 %>%
  select(-moe) %>%
  spread(variable, estimate) %>%
  rename(County = NAME) 

#Can't figure out how to include in the pipe.
names(poverty_IA)[3:30] <- v03
colnames(poverty_IA) <- gsub("!!", "_", colnames(poverty_IA)) 
colnames(poverty_IA) <- gsub(" ", "_", colnames(poverty_IA)) 


##Do not use plyr!!! no round_any()
poverty_IA <- poverty_IA %>%
  group_by(County) %>%
  mutate(Total_Minors = sum(Male_Under_5_years, Male_5_years, Male_6_to_11_years, Male_12_to_14_years, Male_15_years, Male_16_and_17_years, Female_Under_5_years, Female_5_years, Female_6_to_11_years, Female_12_to_14_years, Female_15_years, Female_16_and_17_years), 
         Total_Seniors = sum(Male_65_to_74_years, Female_65_to_74_years, Male_75_years_and_over, Female_75_years_and_over))

poverty_IA$per_minors <- (poverty_IA$Total_Minors / Total_Age_IA$Under_18_years * 100) 
poverty_IA$per_seniors <- (poverty_IA$Total_Seniors/Total_Age_IA$Over_65_years * 100)


poverty_IA <- poverty_IA %>% 
  mutate(iper_minors = cut(per_minors, breaks = c(0, 5, 10, 15, 20, 25, 30, 35),
                            labels = c("<5", "5-10", "10-15", "15-20", "20-25", "25-30", "30-35")),
         iper_seniors = cut(per_minors, breaks = c(0, 5, 10, 15, 20, 25, 30, 35),
                            labels = c("<5", "5-10", "10-15", "15-20", "20-25", "25-30", "30-35"))) 
           

        

#Error: Evaluation error: Specify exactly one of n and width. ?? 
##Error: Evaluation error: 'breaks' are not unique ??
  mutate(Minorsf = cut_interval(Total_Minors, legnth = 3 ))
```

```{r Minors in poverty expl/ join multiplt tables}
poverty_IA %>%
  select(County, per_minors) %>% 
  View()

C1 <- cbind(County = acs_IA$County, acs_IA$total_population, acs_IA$med_family_income, poverty_IA$per_minors, acs_IA$geometry) %>%
  as_tibble() %>%
  rename(total_population = V2, med_family_income = V3, per_minors = V4, geometry = V5) %>%
  mutate(as.numeric(med_family_income))

View(C1)
class(C1$med_family_income)
#cbind converts columns to list, can't use as tibble, so no summarizing or plotting. Use join instead (?)

plot(x = acs_IA$med_family_income, y = poverty_IA$per_minors) 
#This plot is a good step, There is a negative correlation with Minors living in poverty and Median family income with the most variability around $50,000 income level 
names(acs_IA)
A1 <- acs_IA %>%
  select(County, total_population, med_family_income, rounded_income, GEOID, geometry)

names(poverty_IA)
P1 <- poverty_IA %>%
  select(Total_Minors, Total_Seniors, per_minors, GEOID)

names(Total_Age_IA)
T1 <- Total_Age_IA %>%
  select(Under_18_years, Over_65_years, GEOID)

plyr::join(x, y, by = NULL, type = "left", match = "all")
I1 <- plyr::join(x = A1, y = P1, by = "GEOID", type = "full")
#Error in join(x = A1, y = P1, by = GEOID, type = "full") : 
##  could not find function "join".
#I don't want to use plyr, last time it messed my code. =/
```


Van Buren County (with highest farmer to med income) has highest minors living in poverty, nearly 27%.
Cedar county has lowest, 2.5%. Appanoose County has lowest median family income and Dallas county has the highest median income. neither max or min correspond of max or min of minors living in poverty.


```{r Minors poverty graph}

poverty_IA %>%
  ggplot() +
  geom_sf(aes(fill = factor(iper_minors))) + 
  labs(title = "Percentage of Minors Living in Poverty in Iowa", fill = "Minors (%)") + 
  coord_sf(datum = NA) + 
  theme_minimal() +  
  scale_fill_brewer(palette = "OrRd") 

sum(poverty_IA$Total_Minors)/sum(Total_Age_IA$Under_18_years)
```
This graph is a good start, but really we want proportions. Get ratios based on total respective age groups. Percent Minors is based on proportions of minors in poverty vs total minors.A minor is defined as a person under 18 years old. 14.5% of minors in Iowa are living in poverty.Southern Iowa has the highest concentration of minors living in poverty. Is high minor poverty linked to county income or family size? Also look at poverty level of seniors >65years.

```{r Seniors poverty graph}

poverty_IA %>%
  ggplot() +
  geom_sf(aes(fill = factor(iper_seniors))) + 
  labs(title = "Percentage of Seniors Living in Poverty in Iowa", fill = "Seniors (%)") + 
  coord_sf(datum = NA) + 
  theme_minimal() +
  scale_fill_brewer(palette = "OrRd") 

sum(poverty_IA$Total_Seniors)/sum(Total_Age_IA$Over_65_years)
```

```{r Correlation graphs}
#Minors to seniors livngin in poverty, poor correlation
ggplot(data = poverty_IA, aes( x = per_seniors, y = per_minors)) +
  geom_point()
#Minors to median income, strong neg correlation
plot(acs_IA$med_family_income, poverty_IA$per_minors)
#Seniors to median income, weak neg correlation
plot(acs_IA$med_family_income, poverty_IA$per_seniors)


poverty_IA %>%
  summarize( Mean_Minors = mean(per_minors), Med_Minors = median(per_minors),
            Med_Seniors = median(per_seniors), Mean_Seniors = mean(per_seniors)) %>%
  View()
Mean_Minors <- mean(poverty_IA$per_minors) #14.73%
Med_Minors <- median(poverty_IA$per_minors) #13.95
Med_Seniors <- median(poverty_IA$per_seniors) #7.39%
Mean_Seniors <- mean(poverty_IA$per_seniors) #7.31%
```


```{r IA census Median Income graphs}

acs_IA %>%
  ggplot() +
  geom_sf(aes(fill = factor(rounded_income))) + 
  labs(title = "Iowan Median Family Income") + 
  coord_sf(datum = NA) + 
  theme_minimal() 

#check, no more na income values... I wws using a grandparents household income
acs_IA %>%
  count(is.na(med_family_income))
```

```{r}
ggplot(data = acs_IA) +
  geom_sf(aes(fill = Age)) + 
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Iowan Median Age") + 
  coord_sf(datum = NA) + 
  theme_minimal()

#not thrilled about the color, add a legend title. Not terrible interesting as 25 to 40 is not a large span, looking more indepth would be more interesting. Which counties has the most kids (<18yrs) which county is the oldest?
```

```{r Total Iowan farmers}

total_farmers <- sum(acs_IA$total_farmers)
IA_pop <- sum(acs_IA$total_population)
farmers_per <- round(total_farmers / IA_pop * 100, 2)
farmers_per
```
4.85% agrees with the 2017 Census of Agiculture (4.54%)
According to the 5 year acs 2017 study only 4.85% of Iowans are farmers, this shoudl include non-full time farmers. The number seems very low to me, however I wonder this this number includes people who work on farms temporarily or seaonally and if it includes any immigrant seasonal labor. According to a 2016 USDA ag study, in the US only 49.1% of agicultural workers are US citizens. Are these two studies comapring the same data and representing farmers "accurately"

```{r percent farmers in iowa}
#Example of what was previously done... can we manually set cut intervals? 
##census_IAf$per_cFarmer <- cut_interval(census_IAf$per_Farmer, n = 5)

acs_IA %>%
ggplot() +
  geom_sf(aes(fill = factor(farmersf))) + 
  labs(title = "Percentage of working Iowan Farmers", fill = "Percent Farmers") + 
  coord_sf(datum = NA) +
  theme_minimal() + 
  scale_fill_brewer(palette = "Greens") 

```
# error:   mutate(farmersf = cut_interval(farmers_per, n = 5, breaks = c(2.5, 5, 10, 15, Inf), labels = c("<2.5", "2.5-5", "5-10", "10-15", ">15"))) Gave: Evaluation error: length of 'breaks' and labels' differ. ??



```{r farmer income}
#cut_width spaces each bin out by a specific width, like .5 units away. But cannot specify labels in argurement. lables = c("<.25", ".25-.75", ".75-1.25", "2.25-2.75"))
acs_IA %>%
  ggplot() +
  geom_sf(aes(fill = factor(farmersi))) + 
  labs(title = "Ratio of Total Median Income to Median Farming Income for Iowa", fill = "Ratio of Incomes") + 
  coord_sf(datum = NA) + 
  theme_minimal() 
```

```{r Non-highschool graduates}

acs_IA %>%
  summarize(Total_Non_grauates = sum(total_nonHigh) / sum(total_population)*100, 
         With_broadband = mean(nonHigh_computer_broadband_per), 
         With_dialup = mean(nonHigh_computer_dialup_per), 
         With_computer = mean(nonHigh_computer_per), 
         No_computer = mean(nonHigh_no_computer_per)) %>%
  select(-geometry) %>%
  mutate_if(is.numeric, round, 2) %>%
     knitr::kable() %>%
  kableExtra::kable_styling("striped")
```


I misrepresented the non-highschool graduates variable. From tabele: Educational Attainment by Presence of a Computer and Types of Internet Subscription in Household, variables actually describles Highschool equivalent or less. I will remove these variables. 

